Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-02-07 ms/ms.tex
still working ...

One of the most difficult parts of model comparison is finding
parameter sets that are commensurate with many different model
structures. For the most part, our models are too complex to easily
derive analytical correspondences among them. Given a numerical
criterion, such as $r$ (initial exponential growth rate) or $\rzero$ 
(intrinsic reproductive number), we can adjust one or more
parameters by brute force to ensure that all of the models match
according to that criterion. While $\rzero$ is often considered
the most fundamental property of an epidemic, and might thus seem to
be a natural matching criterion, here we focus on matching the initial
growth rate $r$ for several reasons. First, our primary interest is in
the transient evolutionary dynamics of virulence, which are more
strongly affected by $r$ than $\rzero$. Second, $r$ is 
more directly observable in real epidemics; $r$ can be estimated by
fitting an exponential curve to the initial incidence or
prevalence curves \cite{ma_estimating_2014}, while $\rzero$
typically requires either (1) knowledge of \emph{all} epidemic
parameters or (2) calculations based on
$r$ and knowledge of the serial interval or generation interval of the
disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
every run has the same initial exponential growth rate in 
disease prevalence.

@@

For parameters involved in partnership turnover ($c$ and $\rho$), main distinction is between models with and without extra-pair contact. With extra-pair contact, we get gradual decrease in peak time, maximum \Lspvl, and equilibrium \Lspvl\ with increasing turnover rates; Otherwise we get increase in summary statistics.
Although we have not verified this idea analytically, we speculate that increase in turnover rates diminishes the effect of epc (equivalently increasing relative importance of within-partnership contact), therefore selecting for lower \Lspvl.
For models without epc, increased rates of partnership change decrease the level of structured mixing (mimicing epc models), resulting in selection for higher \Lspvl.

@@

Surprisingly, the biggest differences are not between the simplest
and the most complex contact structures, but between the
complex contact structures and models of \emph{intermediate}
complexity. These intermediate-complexity models attempt,
quite reasonably, to add at least some of the realism of
human sexual behaviour, but leave out apparently insignificant
details such as extra-pair contact. If partial complexity
may lead to such mistakes, how can modelers do anything but always
strive to build the most realistic models possible?

>>>>

One of the most difficult parts of model comparison is finding
parameter sets that are commensurate across radically different model
structures. For the most part, our models are too complex to
derive analytical correspondences among the parameters
for different models. Given a numerical
criterion, such as $r$ (initial exponential growth rate) or $\rzero$ 
(intrinsic reproductive number), we can adjust one or more
parameters by brute force to ensure that all of the models match
according to that criterion. While $\rzero$ is often considered
the most fundamental property of an epidemic, and might thus seem to
be a natural matching criterion, here we focus on matching the initial
growth rate $r$ for several reasons. First, our primary interest is in
the transient evolutionary dynamics of virulence, which are more
strongly affected by $r$ than $\rzero$. Second, $r$ is 
more directly observable in real epidemics; $r$ can be estimated by
fitting an exponential curve to the initial incidence or
prevalence curves \cite{ma_estimating_2014}, while $\rzero$
typically requires either (1) knowledge of \emph{all} epidemic
parameters or (2) calculations based on
$r$ and knowledge of the serial interval or generation interval of the
disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
every run has the same initial exponential growth rate in 
disease prevalence.

@@

For parameters involved in partnership turnover ($c$ and $\rho$), the main distinction is between models with and without extra-pair contact. With extra-pair contact, we see a gradual decrease in peak time, maximum \Lspvl, and equilibrium \Lspvl\ with increasing turnover rates.
Increases in the other parameters lead to increases in all
three summary statistics.
We interpret these patterns as evidence
that the increase in turnover rates diminishes the effect of extra-pair contact (equivalently increasing relative importance of within-partnership contact), therefore selecting for lower \Lspvl.
For models without extra-pair contact, increased rates of partnership change decrease the level of structured mixing (mimicing extra-pair contact models), resulting in selection for higher \Lspvl.

@@

The restriction of transmission within stable partnerships
strongly limits eco-evolutionary dynamics by
limiting the maximum speed of epidemic growth. An HIV genotype
that optimizes SPVL to maximize the speed of spread in
a homogeneous population will be sub-optimal in a context
where infection can only spread beyond a partnership once
it dissolves. This finding echoes a long line of
studies that show that population structure leads to the
evolution of ``prudent'' parasites, although most of these studies
focus on equilibrium optima rather than eco-evolutionary dynamics
\citep{herre_factors_1995,kerr_local_2006,webb_theoretical_2013,berngruber_spatial_2015}. 
The more complex contact structures we modeled
mitigate these constraints by allowing HIV to spread among uncoupled individuals (through finite pair-formation) and members of stable partnerships
(through extra-pair contact), albeit at lower rates than within
partnerships.

Thus, we see the biggest differences not between the simplest
and the most complex contact structures, which either
ignore pair structure completely or allow for extra-pair
contact that reduces its impact, but between the
complex contact structures and models of \emph{intermediate}
complexity. These intermediate-complexity models attempt,
quite reasonably, to add at least some of the realism of
human sexual behaviour, but err by neglecting the apparently insignificant
detail of extra-pair contact. If partial complexity
may lead to such mistakes, how can modelers do anything but always
strive to build the most realistic models possible?
